---
title: "No forest is an island"
author: "Maxime Jaunatre"
date: "`r Sys.Date()`"
output: 
    github_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document is set to test the different migration or mosaic setups and develop code to simulate this. This is purely experimental and rely on the dev branch of the `{matreex}` package.

```{r library}
devtools::load_all()
library(dplyr)
library(ggplot2)
options(W_matreex_edist = FALSE)
```

## Regional pool

The idea is to add recruitment from a regional pool with a constant proportion of species and add the recruit produced with a migration rate.

The recruitment function for a species $s$ is edited such that :

$$
R_s = \alpha BA_s^\beta \cdot e^{-\delta BA_s} \cdot e^{-\epsilon BA_h}\\
R_s = [(1-m)\alpha BA_s^\beta +  m \alpha BA_r^\beta] \cdot e^{-\delta BA_s} \cdot e^{-\epsilon BA_h}
$$

$BA_s$ is the local basal area of the species, $BA_r$ is the regional basal area of the species and $BA_h$ is the local basal area of other species. $\alpha$, $\beta$, $\delta$ and $\epsilon$ are parameters fitted for each species. $m \in R: 0 \leq m \leq 1$ 

To simplify, the recruitment is impacted by regional fecundity, but the competition is still local.

### Implementation

This is just the creation of the species required for latter simulations, not much fun yet.

```{r common_value, cache=TRUE}
data("fit_Abies_alba")
data("climate_species")
climate <- subset(climate_species, N == 2 & sp == "Abies_alba", select = -sp)
climate

Abies_ipm <- make_IPM(
    species = "Abies_alba",
    climate = climate,
    fit = fit_Abies_alba,
    clim_lab = "optimum clim",
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Abies_alba) * 1.1),
    BA = 0:120,
    verbose = TRUE
)
Abies_sp <- species(IPM = Abies_ipm, init_pop = def_initBA(30))
Abies_for <- forest(species = list(Abies = Abies_sp))
set.seed(42) # The seed is here for initial population random functions.
Abies_sim <- sim_deter_forest(
    Abies_for,
    tlim = 2000,
    equil_time = 2000, equil_dist = 1, equil_diff = 1,
    SurfEch = 0.03,
    verbose = TRUE
)

data("fit_Fagus_sylvatica")
Fagus_ipm <- make_IPM(
    species = "Fagus_sylvatica",
    climate = climate,
    fit = fit_Fagus_sylvatica,
    clim_lab = "optimum clim",
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
    BA = 0:120, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
Fagus_sp <- species(IPM = Fagus_ipm, init_pop = def_initBA(30))
AbFa_for <- forest(species = list(Abies = Abies_sp,  Fagus = Fagus_sp))
set.seed(42) # The seed is here for initial population random functions.
AbFa_sim <- sim_deter_forest(
    AbFa_for,
    tlim = 2000,
    equil_time = 2000, equil_dist = 1, equil_diff = 1,
    SurfEch = 0.03,
    verbose = TRUE
)
```

Now forest class require to have regional abundance, that is basal area and migration rate.
These values are required to follow the order of the species.

```{r edit_forest}
### regional forest class ####
#' species are initiated with their own functions and only the regional abundance
#' is set in the forest object as well as the migration rate

#' Constructor for forest class
#'
#' Only used in the matreex package
#'
#' @param species List of species created with matreex package.
#' @param harv_rules Vector for harvest rules at the scale of the forest.
#' \describe{
#'   \item{Pmax}{maximum proportion of BAcut / BA}
#'   \item{dBAmin}{the minimum BA to perform cut}
#'   \item{freq}{Frequence at which the harvest will be executed.}
#' }
#' @param regional_abundance list of vector with size distribution for each species.
#' This is not a direct species relative abundance, but I don't know how to implement this...help ,
#' @param migration_rate numeric vector with a migration rate in percentage between 1 et and 0.
#'
#' @importFrom purrr map_chr
#'
#' @keywords internal
#' @export
new_forest <- function(species = list(),
                       harv_rules = c(Pmax = 0.25, dBAmin = 3, freq = 1, alpha = 1),
                       regional_abundance = NULL,
                       migration_rate = NULL
){

    sp <- map_chr(species, sp_name)
    names(species) <- sp
    if(!is_null(regional_abundance)){
        names(regional_abundance) <- sp
        names(migration_rate) <- sp
    }
    forest <- list(
        species = species, harv_rules = harv_rules,
        info = list(species = sp,
                    clim_lab = map_chr(species, climatic)),
        regional_abundance = regional_abundance,
        migration_rate = migration_rate
    )

    if(!is_null(regional_abundance)){
        class(forest) <- c("forest", "reg_forest")
    } else {
        class(forest) <- "forest"
    }

    return(forest)
}

#' validator for forest class.
#'
#' @param x forest class object
#'
#' @import checkmate
#'
#' @noRd
validate_forest <- function(x){

    regional <- inherits(x, "reg_forest")
    values <- unclass(x)
    names <- attr(x, "names")

    map(values$species, validate_species)
    # TODO check forest harv rules

    clim_lab <- values$info$clim_lab
    if(length(unique(clim_lab)) > 1){
        clim_ipm <- clim_lab[clim_lab != "mu_gr"]
        if(length(clim_ipm) > 1){ # D & F
            stop(paste0("Some ipm species are not defined with the same climatic name.",
                        "Check it with : map_chr(species, climatic)"))
        }
    }

    # check the regional pool settings
    if(regional){

        assertNumeric(values$migration_rate, len = length(values$species),
                      lower = 0, upper = 1)
        if(all(values$migration_rate == 0)){
            warning("All migration rate are 0, the regional pool of this forest is deleted")
            x$regional_abundance <- NULL
            x$migration_rate <- NULL
            class(x) <- "forest"

            return(invisible(x))
        }

        assertSubset(names(values$migration_rate), names(values$species))
        # length_meshs <- map_dbl(values$species, ~ length(.x$IPM$mesh))

        # assertList(values$regional_abundance, types = "numeric",
                   # len = length(values$species))
        # if(any(lengths(values$regional_abundance) != length_meshs)){
            # stop("regional abundance numeric vector should be the length of the species mesh.")
        # }

        assertSubset(names(values$regional_abundance), names(values$species))
    }

    invisible(x)
}

#' Create a new forest for simulation
#'
#' A forest is a group of one of multiple species to silumate along time using
#' the IPM defined for each species and harvest rules.
#'
#' @inheritParams new_forest
#'
#' @export
forest <- function(species = list(),
                   harv_rules = c(Pmax = 0.25, dBAmin = 3, freq = 1, alpha = 1),
                   regional_abundance = NULL,
                   migration_rate = NULL
){

    res <- validate_forest(new_forest(
        species = species,
        harv_rules = harv_rules,
        regional_abundance = regional_abundance,
        migration_rate = migration_rate
    ))

    return(res)
}
```


Another key edit is that we need to create an alternative recruitment function where the fecundity rely on the regional basal area. The default recruitment function is $rec\_fun(BA_s, BA_h)$ and the new one is $rec\_fun_{regional}(BA_s, BA_r, BA_h)$ : 

$$
R_s = (1-m) \cdot rec\_fun(BA_s, BA_h) + m \cdot rec\_fun_{regional}(BA_s, BA_r, BA_h) \\
R_s = (1-m)\alpha BA_s^\beta \cdot e^{-\delta BA_s} \cdot e^{-\epsilon BA_h} + m \alpha BA_r^\beta \cdot e^{-\delta BA_s} \cdot e^{-\epsilon BA_h}
$$

This edit is possible with the `regional` option of the internal function `exp_recFun()`.

```{r edit_recfun}
list_covs <- climate
params <- fit_Abies_alba$rec$params_m
matreex:::exp_recFun(params, list_covs)
matreex:::exp_recFun(params, list_covs, regional = TRUE)
```

*This edit highlight that our simulation are heavily dependant on the basal area for competition and not very open for other competition indices.

### Simulations

Now are several examples to test if it works.

First, I start from equilibrium and compare with and without migration. I expect no differences.

```{r onespsim}
equil_dist <- dplyr::filter(Abies_sim, equil, var == "n") %>% dplyr::pull(value)
equil_BA <- dplyr::filter(Abies_sim, equil, var == "BAsp") %>% dplyr::pull(value)


Abies_spE <- Abies_sp
Abies_spE$init_pop <- def_init_k(equil_dist)
comp_time <- 500

regional_Abies <- forest(
    species = list(Abies = Abies_spE),
    regional_abundance = equil_BA,
    migration_rate = c(Abies = 0.05)
)

regional_sim <- sim_deter_forest(
    regional_Abies,
    tlim = comp_time,
    equil_time = comp_time, equil_dist = 10,
    SurfEch = 0.03,
    verbose = TRUE
)

isol_sim <- sim_deter_forest(
    forest(
        species = list(Abies = Abies_spE),
        regional_abundance = list(Abies = equil_dist),
        migration_rate = c(Abies = 0)
    ),
    tlim = comp_time,
    equil_time = comp_time, equil_dist = 10,
    SurfEch = 0.03,
    verbose = TRUE
)


regional_pool <- dplyr::bind_rows(regional = regional_sim, isolation = isol_sim,
                                  .id = "migration")

regional_pool %>%
    dplyr::filter(var %in% c("BAsp", "N"), !equil) %>%
    ggplot(aes(x = time, y = value, color = migration, linetype = migration)) +
    geom_line(linewidth = 1) +
    facet_wrap(~ var, scales =  "free_y") +
    NULL
```

Now we test two species. I expect no variations from equilibrium because we start from equilibrium and the regional pool is at equilibrium.

```{r twosp_sim}
e2_dist <- dplyr::filter(AbFa_sim, equil, var == "n") %>%
    dplyr::group_by(species) %>%
    dplyr::group_split() %>% map(pull, value)
e2_BA <- dplyr::filter(AbFa_sim, equil, var == "BAsp") %>% dplyr::pull(value)


Abies_spEn <- Abies_sp
Abies_spEn$init_pop <- def_init_k(e2_dist[[1]])
Fagus_spEn <- Fagus_sp
Fagus_spEn$init_pop <- def_init_k(e2_dist[[2]])


regional_AbFa <- forest(
    species = list(Abies = Abies_spEn, Fagus = Fagus_spEn),
    regional_abundance = e2_BA,
    migration_rate = c(Abies = 0.1, Fagus = 0.1)
)

sp2_sim <- sim_deter_forest(
    regional_AbFa,
    tlim = comp_time,
    equil_time = comp_time, equil_dist = 10,
    SurfEch = 0.03,
    verbose = TRUE
)

sp2_sim %>%
    dplyr::filter(var %in% c("BAsp", "N"), !equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    geom_line(linewidth = .4) +
    facet_wrap(~ var, scales =  "free_y") +
    NULL
```

One species is absent and colonise from regional pool. **Note : starting a simulation with a species with 0 effectif is generally a bad idea, now it's possible but do NOT abuse this possibility !**

I expect the new species to grow, but the lag will be visible. I don't simulate until equilibrium, but I guess it's reached at some point.

```{r}
Fagus_spE0 <- Fagus_sp
Fagus_spE0$init_pop <- def_init_k(e2_dist[[2]] * 0)


invasive_AbFa <- forest(
    species = list(Abies = Abies_spE, Fagus = Fagus_spE0),
    regional_abundance = e2_BA,
    migration_rate = c(Abies = 0.1, Fagus = 0.1)
)

invas_sim <- sim_deter_forest(
    invasive_AbFa,
    tlim = comp_time,
    equil_time = comp_time, equil_dist = 10,
    SurfEch = 0.03,
    verbose = TRUE
)

invas_sim %>%
    dplyr::filter(var %in% c("BAsp", "N"), !equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    geom_line(linewidth = .4) +
    facet_wrap(~ var, scales =  "free_y") +
    NULL
```

With the same start as those I used when looking from equilibrium, I expect to reach equilibrium faster. This is because the recruitment is drived by the regional pool. However, this speed might be dependant on the migration rate.

```{r}
rand_AbFa <- forest(species = list(Abies = Abies_sp,  Fagus = Fagus_sp),
                    regional_abundance = e2_BA,
                    migration_rate = c(Abies = 0.5, Fagus = 0.5)
)

set.seed(42)
rand_sim <- sim_deter_forest(
    rand_AbFa,
    tlim = 2000,
    equil_time = 2000, equil_dist = 10,
    SurfEch = 0.03,
    verbose = TRUE
)

works <- dplyr::bind_rows(regional = rand_sim, isolation = AbFa_sim,
                 .id = "migration")

works %>%
    dplyr::filter(var %in% c("BAsp", "N"), !equil) %>%
    ggplot(aes(x = time, y = value, color = species, linetype = migration)) +
    geom_line(linewidth = .4) +
    facet_wrap(~ var, scales =  "free_y") +
    NULL
```


