---
title: "Climate variation simulation"
output: pdf_document
date: "2022-11-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Testing mu simulation

This document show how to simulate a species with mu growth pre-computation and not integrate a classical IPM.
For now, the package is in experimental state and use the branch *fast_int*. For this example we will use the species *Picea abies* as an example.

```{r constant}
# Libraries
library(ggplot2)
library(dplyr)
library(devtools)

# Loading all functions of the package
devtools::load_all()
species <- "Picea_abies"
data(list = paste0("fit_", species))
fit <- eval(parse(text=paste0("fit_", species)))
climate <- subset(climate_species, sp == species, select = -sp)
```

## Get Mu values

Mu is an intermediate value computed during the simulation.

The first step consist in computing all mu values for a species in its climatic and basal area range. For this, we compute the mu range of the species and simulate a matrix for each values of mu, with a small step. Keep in mind that the time of this computation rely mostly on the step and this step will influence the precision of the simulation later. The mu range step is done inside the `make_mu_gr` function call, but it's possible to check it before.

```{r range_mu}
mu_range <- getRangemu(
    climate = climate, fit = fit, BA = seq(0, 200, by = 10), 
    mesh = seq(90, get_maxdbh(fit_Picea_abies) * 1.1, by = 2))
mu_range
```

*Note : Low values highly depend on maximum basal area, so, one can limit the mu computation time if he expect the basal area to stay low during simulation.*

Making the mu matrix takes more or less the same arguments as the function `make_IPM` in term of integration levels for Gauss-Legendre and midbin methods. The important argument that limit computation time is stepMu.

```{r make_mu}
mu_Picea_abies <- make_mu_gr(
    species = "Picea_abies", fit = fit_Picea_abies,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    verbose = TRUE, stepMu = 0.001)
```

This object is a matrix with as many row as integrated cells for a given mu, and a range of mu in column. During simulation, the loop will compute the mu needed for climate and basal area for a given time and extract from the matrix the required columns. 

**This step is 5 times longer than extracting the IPM matrix, this is the bottleneck during simulation.**

```{r get step_IPM}
step_climate <- subset(climate, N ==2, select = -N)
step_climate <- drop(as.matrix(step_climate))
Picea_IPM_BA20 <- get_step_IPM(x = mu_Picea_abies, BA = 20, 
                               climate = step_climate, sim_corr = "cut")
Picea_IPM_BA20[12:17, 12:17]
dim(Picea_IPM_BA20)
```

## Simulations

Once the mu matrix is created, creating species and simulating is really close to simulating with IPM. Because the mu matrix are not defined for a specific climate, we need to set one to simulate on it. For now we only set a single climate that will not change during simulation.

```{r simulation}
time <- 1000

Picea_abies <- species(IPM = mu_Picea_abies, init_pop = def_initBA(20), 
                       harvest_fun = def_harv)

forest <- forest(species = list(mu_Picea = Picea_abies))
set.seed(42)
memor_mu <- sim_deter_forest.forest(forest, tlim = time, climate = step_climate,
                                    equil_dist = time, equil_time = time,
                                    verbose = TRUE, correction = "cut")

memor_mu %>%
    filter(var %in% c("BAsp", "H", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4) +
    NULL
```

We can compare this with a classical IPM simulation to see that it's not so different. Obviously this takes some times since integration takes longer. To limit this, we only integrate for basal area up to 100.

```{r compare_IPM}
ipm_Picea <- make_IPM(
    "Picea_abies", step_climate, "opt_Picab_clim", fit = fit_Picea_abies,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:100, verbose = TRUE
)
Picea_abies_ipm <- species(IPM = ipm_Picea, init_pop = def_initBA(20), 
                           harvest_fun = def_harv)
forest_ipm <- forest(species = list(ipm_Picea = Picea_abies_ipm))
set.seed(42)
memor_ipm <- sim_deter_forest.forest(forest_ipm, tlim = time, 
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut")
```

Below is a plot that combine both simulations and compute the difference along time.

```{r compare_plot}
e_memor <- dplyr::bind_rows(ipm = memor_ipm, mu = memor_mu, .id = "meth")
e_memor %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = meth)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(size = .4, linetype = "dotted") + geom_point(size = .4) +
    NULL

e_memor %>%
    filter(var %in% c("BAsp"), ! equil) %>%
    group_by(time) %>% summarise(value = diff(value)) %>%
    ggplot(aes(x = time, y = value)) +
    ylab("BA difference") +
    geom_line(linewidth = .4, linetype = "dotted") + geom_point(size = .4) +
    NULL
```

## Climate evolution in time

Previous simulation specified a single climate that is used from start to finish, with only a variation on BA. The interest of using precomputed values is too create new IPM matrix for a lot more climates. But as integrating an IPM takes around 2 or minutes, integrating a climate gradient would takes way too long, and only to use few basal area values when we integrate between 0 and 200.

Of course we now need to define a climate at each time step of the future simulation. For example, the simulation will start on cold edge and migrate to hot edge in a thousand years, with a small random effect.

**This code is an example and some functions may be added to the package later.**

```{r future_in_pkg}
# future function for the package matreex. 
# I may need some help on what's needed or usefull

#' @param x a named values for a climatic variable. 
#' @param inv_null If TRUE, the inverse variable is \code{varb = 1/(var+1)}, in
#' case of var can take 0 for value.
#' Example : \code{x = c(sgdd = 2000)}
value2sqb <- function(x, inv_null = FALSE){
    
    nms <- names(x)
    res <- numeric(3)
    names(res) <- c(nms, paste0(nms, c("2", "b")))
    res[nms] <- x
    res[paste0(nms, "2")] <- x^2
    res[paste0(nms, "b")] <- 1/(x + inv_null)
    
    res
}

#' @param climate Named vector of climatic variables.
#' @param inv_null If TRUE, the inverse variable is \code{varb = 1/(var+1)}, in
#' case of var can take 0 for value.
#' Example : \code {climate = c(sgdd = 2000, wai = 0.16) ; in_null = c(sgdd = FALSE, wai = TRUE)}
expand_clim <- function(climate, inv_null){
    
    assertNumeric(climate, any.missing = FALSE)
    assertLogical(inv_null, any.missing = FALSE)
    
    nms <- names(climate)
    res <- vector("list", length(climate))
    for(x in seq_along(nms)){
        res[[x]] <- value2sqb(climate[nms[x]], inv_null[nms[x]])
    }
    res <- unlist(res)
    
    return(res)
}
# May be usefull to create the same function but to work on data.frame 
# or matrices likes below code
```


```{r build_climate}
n <- time
tmp <- subset(climate, N != 2, select = c(sgdd,wai, N))

wai <- seq(tmp$wai[2], tmp$wai[1], length.out = n) + runif(n, -0.05, 0.05)
wai <- pmax(wai, 0) ; wai <- pmin(wai, 1) # range the values after runif
sim_clim <- as.matrix(data.frame(
    sgdd = seq(tmp$sgdd[2], tmp$sgdd[1], length.out = n) + runif(n, -50, 50),
    wai = wai, 
    t = 1:n
))
# We need to specify if variables can be set 0 for variable compution of
# varb = 1 / var. If true varb = 1 / (var + 1)
inv_null <- c(sgdd = FALSE, wai = TRUE)

exp_clim <- vector("list", n)
for(i in 1:n){
    exp_clim[[i]] <- c(expand_clim(sim_clim[i, -3], inv_null), t = i)
}
exp_clim <- do.call("rbind", exp_clim)
dim(exp_clim)
exp_clim[c(1:3, 997:1000),]
```

Once the climate is set, we can simulate a species. I recreate the species to start with a size distribution closer to the equilibrium reached in previous simulations. Don't forget to modify this distribution with the surface sampled (`SurfEch`).

```{r clim_sim}
distrib <- filter(memor_mu, equil, var == "m") %>% pull(value) * 0.03
Picea_abies <- species(IPM = mu_Picea_abies, init_pop = def_init_k(distrib), 
                       harvest_fun = def_harv)
clim_forest <- forest(species = list(clim_Picea = Picea_abies))
set.seed(42)
memor_clim <- sim_deter_forest.forest(clim_forest, tlim = time, climate = exp_clim,
                                      equil_dist = time, equil_time = time,
                                      verbose = TRUE, correction = "cut")

memor_clim %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4) +
    NULL
```

I think this fast decay at start is because the distribution comes from an equilibrium for the "optimal" climate and not the the cold edge, but past this point an increase in *sgdd* and *wai* seems to increase basal area along with a decrease of population size. However, the climate variation is relatively quick and we may need to test more hypothesis. Lastly, the evolution of metrics is not smooth, so the random effect on climate may be effective.

## Testing effect of stepMu

The BA difference is already pretty low between IPM and mu simulations, but this section will illustrate that it's directly linked with `stepMu` value. Error is negligeable for stepMu of 0.0001, but computing mu values takes around 2 minutes.

```{r stepMu}
steps <- c(0.05, 0.01, 0.005, 0.001, 0.0005)
diff_list <- vector("list", length(steps))

for(s in seq_along(steps)){
    cat(sprintf(" %.4f", steps[s]))
    mu_Picea <- make_mu_gr(
        species = "Picea_abies", fit = fit_Picea_abies,
        mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
        stepMu = steps[s])
    
    Picea <- species(IPM = mu_Picea, init_pop = def_initBA(20), 
                     harvest_fun = def_harv)
    
    forest <- forest(species = list(Picea = Picea))
    set.seed(42)
    memor <- sim_deter_forest.forest(forest, tlim = time, 
                                     climate = step_climate,
                                     equil_dist = time, equil_time = time,
                                     correction = "cut") 
    
    diff_list[[s]] <- dplyr::bind_rows(ipm = memor_ipm, mu = memor, 
                                       .id = "meth") %>%
        filter(var %in% c("BAsp"), ! equil) %>%
        group_by(time) %>% summarise(value = diff(value))
}

names(diff_list) <- c("a: 5e-2", "b: 1e-2", "c: 5e-3", "d: 1e-3", "e: 5e-4")
res <- do.call("bind_rows", c(diff_list, list(.id = "stepMu")))
res %>%
    ggplot(aes(x = time, y = value, color = stepMu)) +
    ylab("BA difference") +
    geom_line(linewidth = .4) +
    NULL

res %>% group_by(stepMu) %>%
    summarise(min = min(value), max = max(value), d = max - min)
```


## Multiple species simulations

We had a second species with mu values. This is just so we can have a distribution from species equilibrium at the climate given at very start of this document.

```{r Abies_alba}
mu_Abies_alba <- make_mu_gr(
    species = "Abies_alba", fit = fit_Abies_alba,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Abies_alba) * 1.1),
    verbose = TRUE, stepMu = 0.001)

Abies_alba <- species(IPM = mu_Abies_alba, init_pop = def_initBA(20), 
                      harvest_fun = def_harv)

forest <- forest(species = list(mu_Abies = Abies_alba))
set.seed(42)
memor_abies <- sim_deter_forest.forest(forest, tlim = time, climate = step_climate,
                                       equil_dist = time, equil_time = time,
                                       verbose = TRUE, correction = "cut")

distrib_abies <- filter(memor_abies, equil, var == "m") %>% pull(value) * 0.03
Abies_alba <- species(IPM = mu_Abies_alba, 
                      init_pop = def_init_k(distrib_abies), 
                      harvest_fun = def_harv)
```

We create a forest with 2 species integrated with mu values. From this, we just need to run the simulation. Please note that the climate can be modified along time as before because all species use mu values. Another thing to note is the simulation time that is correlated with the number of mu species used.

```{r}
nsp_forest <- forest(species = list(Picea = Picea_abies, Abies = Abies_alba))
set.seed(42)
memor_nsp <- sim_deter_forest.forest(nsp_forest, tlim = time, climate = exp_clim,
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut") 

memor_nsp %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4) +
    NULL
```

If we want to simulate with species defined from mu and ipm, we can mix them but the climate used to integrate the ipm will be used. Providing a climate is thus useless in this case. 

```{r}
Picea_abies_ipm$init_pop <- def_init_k(distrib)
msp_forest <- forest(species = list(Picea = Picea_abies_ipm, Abies = Abies_alba))
set.seed(42)
memor_msp <- sim_deter_forest.forest(msp_forest, tlim = time, climate = exp_clim,
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut")

memor_msp %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4) +
    NULL
```
