---
title: "Perturbations"
author: "Maxime Jaunatre"
date: "2022-12-08"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Testing simulation with perturbations

This document show how to simulate a species with perturbation occuring in time.
For now, the package is in experimental state and use the branch *tempest*. For this example we will use the species *Picea abies* as an example.

Only the Julien disturbance model is explained in this case.

```{r constant make ipm}
# Libraries
library(ggplot2)
library(dplyr)
library(devtools)

# Loading all functions of the package
devtools::load_all()
species <- "Picea_abies"
data(list = paste0("fit_", species))
fit <- eval(parse(text=paste0("fit_", species)))
climate <- subset(climate_species, sp == species & N == 2, select = -c(N, sp))
climate <- drop(as.matrix(climate))
ipm_Picea <- make_IPM(
    "Picea_abies", climate, "opt_Picab_clim", fit = fit_Picea_abies,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:100, verbose = TRUE
)
Picea_abies <- species(IPM = ipm_Picea, init_pop = def_initBA(40),
                       harvest_fun = def_harv)
```

We want to start our simulations with an equilibrium size distribution so we compute here first

```{r equilibrium}
forest_ipm <- new_forest(species = list(Picea = Picea_abies))
time <- 3000
set.seed(42)
memor <- sim_deter_forest.forest(forest_ipm, tlim = time,
                                     equil_dist = 250, equil_time = time,
                                     verbose = TRUE, correction = "cut") %>%
    tree_format()

memor %>%
    filter(var %in% c("BAsp"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(size = .4) + geom_point(size = .4) +
    NULL

equil <- memor %>%
    filter(var == "m", equil) %>% pull(value)
```


## Disturbance

### Definition of disturbance

We define a disturbance by few parameters used later in the formula.

* $I$ its intensity

* $t$ the duration of the disturbance. Here we need to set it to 5 because it's the time between two census.

* $type$ the class of disturbance. This is often a label in `"storm"`, `"fire"` and `"biotic"`.
This is not used in the formula but to filter species parameters fitted.

We can set all of this in a data.frame object. We had a last column named **IsSurv**. It's used to tell the simulation if the survival part of the IPM is needed during a disturbance. In this case, the data do not allow differentiation between disturbance mortality and background mortality. Therefore, we need to deactivate baseline mortality so that it is not double counted.

```{r}
ex_disturb <- data.frame(type = "storm", intensity = 0.5, duration = 5, IsSurv = FALSE)
```


### Impact on population

The disturbance impact on the population result from parameters computed by Julien. They compose a function that takes the size distribution, quadratic diameter of the species, intensity ($I$) and duration ($t$) of the disturbance. A set of parameters was made for each type of disturbance and species.

$$
dqm = \sqrt{\frac{\sum_{i=1}^m size_i^2 \times value_i}{\sum_{i=1}^m value_i}} \\
logratio = log(\frac{size}{dqm}) \\
dbh.scaled = dbh.intercept + size \times dbh.slope \\
logratio.scaled = logratio.intercept + logratio \times logratio.slope \\
X_{t+1} = X_t \times (1 - plogis(a_0 + a_1 \times logratio.scaled + b \times I^{c \times dbh.scaled}))^t 
$$
The parameters are estimated with Bayesian computations. The mean of all estimations are stored inside the package for each combination of species and disturbance type.

```{r data_pkg}
(coefs <- filter(treeforce::disturb_coef, species == "Picea_abies"))
Picea_abies$disturb_coef <- coefs
```

Linked with this set of parameters, we need to provide a disturbance function to the species we want to simulate. *The species is initiated with an empty function that will throw warnings.*

```{r disturb_fun}
#' Disturbance function
#' 
#' @param x population state distribution at time t
#' @param species The species class object of interest to get mesh and RDIcoef
#' values from. RDIcoef is a one line dataframe with RDI coefficient for one
#' species.
#' @param disturb Disturbance parameters. Highly depend on the disturbance 
#' impact parameters given to the species.
#' @param ... Not used in this case.
#' 
disturb_fun <- function(x, species, disturb = NULL, ...){
    
    size <- species$IPM$mesh
    coef <- species$disturb_coef
    if(any(disturb$type %in% coef$disturbance)){
        coef <- subset(coef, disturbance == disturb$type)
    } else {
        stop(sprintf("The species %s miss this disturbance type (%s) parameters",
                     sp_name(species), disturb$type))
    }
    
    
    qmd <- treeforce::QMD(size = size, n = x)
    logratio <-  log(size / qmd)
    dbh.scaled = coef$dbh.intercept + size * coef$dbh.slope
    logratio.scaled = coef$logratio.intercept + logratio * coef$logratio.slope
    Pkill <- plogis(coef$a0 + coef$a1 * logratio.scaled + 
                        coef$b * disturb$intensity^(coef$c * dbh.scaled)) ^ disturb$duration 
    
    return(x* Pkill) # always return the mortality distribution
}

Picea_abies$disturb_fun <- disturb_fun

plot(equil, type= "l", xlab = "size index", ylab = "n")
lines(1:700, equil - disturb_fun(equil, Picea_abies, ex_disturb), col = "red")
legend("topright", c("Distribution", "After Disturbance"), 
       lty = c(1, 1), col = c(1, 2))
```

## Simulations

```{r note, echo = FALSE}
# We need to remove survival during perturbation. 
# This is done during simulation at the step where we 
# get the IPM matrix with get_step_IPM.
# TODO  find a way to show this !
```

Running a simulation takes the same parameters as usual, with an added data.frame with disturbance along time. *We need to think about a clean way to build this table...*

```{r disturbance table}
time <- 2500
disturb <- data.frame(dist =  FALSE, type = "none", intensity = 0, 
                            duration = 1, IsSurv = TRUE, t = 1:time)

disturb[100, ] <- data.frame(dist = TRUE, type = "storm", intensity = 0.5, 
                             duration = 5, IsSurv = FALSE, t = 100)
disturb[98:102, ]
```



```{r sim}
load_all()
Picea_abies$init_pop <- def_init_k(equil * 0.03)
forest_ipm <- new_forest(species = list(Picea = Picea_abies))
set.seed(42)
memor <- sim_deter_forest.forest(forest_ipm, tlim = time,
                                 equil_dist = time, equil_time = time,
                                 disturbance  = disturb,
                                 verbose = TRUE, correction = "cut") %>%
    tree_format()

memor %>%
    filter(var %in% c("BAsp", "H", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(size = .4) + geom_point(size = .4) +
    NULL
```

## Targets

* Multiple species simulations


