---
title: "Disturbance models"
package: matreex
output: 
    github_document:
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Disturbance models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
# bibliography: references.bib 
link-citations: true

<style>
body {
text-align: justify}
</style>
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(cli.progress_show_after = 600) 
# 10 minutes before showing a cli progress bar
```


This vignette illustrates how to use disturbance scenarii with `{matreex}` package. A disturbance scenario provides the timing and intensity at which some type of disturbance will affect mortality of trees in each size classes. The tree killed by the disturbance are saved within the harvested output.

The basic `{matreex}` functions are shown in [a previous introduction vignette](matreex.html).


# Simulations input

## Define a species

The first step is the IPM integration and species object creation. This part is common with basic usage of the package, so nothing is very important here.

**Please keep in mind this computation is intensive and may take few minutes !**

```{r make_IPM_Picea}
library(matreex)
library(dplyr)
library(ggplot2)

# Loading all functions of the package
devtools::load_all() # TODO remove

# Load fitted model for a species
# fit_species # list of all species in dataset
data("fit_Picea_abies")

# Load associated climate
data("climate_species")
climate <- subset(climate_species, N == 2 & sp == "Picea_abies", select = -sp)
# see ?climate_species to understand the filtering of N.
climate

Picea_ipm <- make_IPM(
    species = "Picea_abies", 
    climate = climate, 
    fit = fit_Picea_abies,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:70, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
```

```{r hide warn, echo=FALSE, }
options(W_matreex_edist = FALSE)
```

Latter simulations will start at equilibrium, so we compute this before anything. `Picea_equil` is the equilibrium distribution after 2500 years.

```{r equilibrium}
Picea_sp <- species(IPM = Picea_ipm, init_pop = def_initBA(40))
Picea_for <- forest(species = list(Picea = Picea_sp))
set.seed(42)
Picea_sim <- sim_deter_forest(Picea_for, tlim = 2500,
                              equil_dist = 250, equil_time = 2500,
                              verbose = TRUE) 

Picea_equil <- filter(Picea_sim, var == "n", equil) %>% pull(value)
```


# Disturbance

## Definition of disturbance

We define a disturbance by few parameters used later in the formula. Details on the model definition and dataset used can be found in @Barrere2023.

* $I$, its intensity between 0 and 1.

* $type$, the class of disturbance. This is a label in `"storm"`, `"fire"` and `"biotic"`.
This is not used in the formula but to filter species parameters fitted.

We can set all of this in a data.frame object. A last column column named **IsSurv** is used to tell the simulation if the survival part of the IPM is needed during a disturbance. In the case of @Barrere2023, the data does not allow differentiation between disturbance mortality and background mortality. Therefore, we need to deactivate baseline mortality so that it is not applied twice.

```{r}
ex_disturb <- data.frame(type = "storm", intensity = 0.5, IsSurv = FALSE)
```


## Impact on population

The disturbance impact on the population result from parameters computed in @Barrere2023. They compose a function that takes the size distribution, quadratic diameter of the species, intensity ($I$) and duration ($t$) of the disturbance. A set of parameters was made for each type of disturbance and species.

$$
dqm = \sqrt{\frac{\sum_{i=1}^m size_i^2 \times value_i}{\sum_{i=1}^m value_i}} \\
logratio = log(\frac{size}{dqm}) \\
dbh.scaled = dbh.intercept + size \times dbh.slope \\
logratio.scaled = logratio.intercept + logratio \times logratio.slope \\
X_{t+1} = X_t \times (1 - plogis(a_0 + a_1 \times logratio.scaled + b \times I^{c \times dbh.scaled}))^t 
$$
The parameters are estimated with Bayesian computations. The mean of all estimations are stored inside the package for each combination of species and disturbance type.

```{r data_pkg}
(Picea_coefs <- filter(matreex::disturb_coef, species == "Picea_abies"))
Picea_sp$disturb_coef <- Picea_coefs
```

Linked with this set of parameters, we need to provide a disturbance function to the species we want to simulate. *The species is initiated with a default empty function that will throw warnings.* The function also require the quadratic mean diameter of the species.

```{r debug_fun, eval = FALSE, echo=FALSE}
x <- Picea_equil
species <- Picea_sp
disturb <- data.frame(type = "storm", intensity = 0.7, IsSurv = FALSE)
qmd <-QMD(Picea_sp$IPM$mesh, x)
```

Below is an example of the difference in size distribution before and after a disturbance of type `"storm"` and intensity `0.7`.

```{r disturb_fun}
#' Disturbance function
matreex::disturb_fun
Picea_sp$disturb_fun <- disturb_fun
```


```{r disturb_ex, echo = FALSE}
ex_disturb <- data.frame(type = "storm", intensity = 0.7, IsSurv = FALSE)
qmd <-QMD(Picea_sp$IPM$mesh, Picea_equil)

plot(Picea_equil, type= "l", xlab = "size index", ylab = "n")
lines(1:(700 + as.numeric(Picea_sp$IPM$info["delay"])), 
      Picea_equil - disturb_fun(Picea_equil, Picea_sp, ex_disturb, qmd = qmd), 
      col = "red")
legend("topright", c("Distribution", "After Disturbance"),
       lty = c(1, 1), col = c(1, 2))
```

Any user can define the `disturb_fun()` function they need and provide parameters for the wanted species. Obviously, this type of modification have not been thoroughly tested and thus is more prone to error so don't hesitate to contact `{matreex}` maintainer in case of trouble.

## Simulations

Running a simulation takes the same parameters as usual, with an added data.frame with disturbance along time. We will start from equilibrium and have a disturbance after 100 years.

```{r disturbance table}
disturb <- data.frame(type = "storm", intensity = 0.2, IsSurv = FALSE, t = 100)
```

**Disturbance size distribution is saved in harvest output. Harvest and basal mortality are canceled when a disturbance happens.**

```{r sim}
Picea_sp$init_pop <- def_init_k(Picea_equil)

disturb_picea <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp)),
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb, # providing disturbance table
    verbose = TRUE)

disturb_picea %>%
    filter(var %in% c("BAsp", "H"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

<!--
## Delay

I test delay here just to be sure. And there was a bug in get_step_IPM so I was right. 

```{r delay, eval = FALSE}
load_all()
time <- 1500
(disturb <- data.frame(type = "storm", intensity = 0.2, 
                       IsSurv = FALSE, t = 600))
forest_d <- forest(species = list(Abies = delay(Abies_alba, 5)))
set.seed(42)
memor_d <- sim_deter_forest.forest(forest_d, tlim = time,
                                 equil_dist = 250, equil_time = time,
                                 disturbance  = disturb,
                                 harvest = "Uneven", targetBA = 30,
                                 verbose = TRUE, correction = "cut")

memor_d %>%
    filter(var %in% c("BAsp", "N"), ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(size = .4) + geom_point(size = .4) +
    NULL
```
-->

## Multispecific simulations 

Multi-specific simulations are performed like the simulations previously illustrated. We also change the harvesting model to illustrate other kind of simulations.

```{r abies_ipm}
data("fit_Fagus_sylvatica")

Fagus_ipm <- make_IPM(
    species = "Fagus_sylvatica", 
    climate = climate, 
    fit = fit_Fagus_sylvatica,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
    BA = 0:70, 
    verbose = TRUE
)
Fagus_sp <- species(
    IPM = Fagus_ipm, init_pop = def_initBA(30),
    disturb_fun = disturb_fun, 
    harvest_fun = Uneven_harv,
    disturb_coef = filter(matreex::disturb_coef, species == "Fagus_sylvatica")
)
```

```{r nsp_sim}
time <- 1500
disturb <- data.frame(type = "storm", intensity = 0.2, IsSurv = FALSE, t = 500)
Picea_sp$harvest_fun <- Uneven_harv

set.seed(42)
disturb_PiFa <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp, Fagus = Fagus_sp),
           harv_rules = c(Pmax = 0.25, dBAmin = 3, freq = 10, alpha = 1)), 
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb,
    harvest = "Uneven", targetBA = 30,
    verbose = TRUE, correction = "cut")

disturb_PiFa %>%
    filter(var %in% c("BAsp", "H"), time > 450 & time < 600, ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```


## Linear stabilizing effect of species mixture

When mixing coniferous and broadleaf species, an additional effect needs to be taken into account based on the survival models by @Brandl2020. This mixture effect is a linear function scaled to the data used to compute disturbance coefficient present in the package (@Barrere2023).

To include this effect, the probability of disturbance mortality is multiplied by the effect and the share of conifers.

$$
f_{mixture.effect}\left(share.conifers\right) = 0.4767 + 0.6015 \cdot share.conifers \\
h_{new}\left(share.conifers\right) = h_{disturb} \cdot f_{mixture.effect}\left(share.conifers\right)
$$

To compute the share of conifers, each species must have a type as shown below. This can be set when creating the species object or edited later :

```{r type}
Picea_sp$info["type"]
Picea_sp$info["type"] <- "Coniferous"
Fagus_sp$info["type"] <- "Broadleaf"
```

An alternative disturbance function is provided by the package as `disturb_fun_mixt()`, to accept the input of the coniferous percentage computed at the plot scale before disturbing each species and to modify the mortality probability (`Pkill`) if the species is a coniferous.

```{r mixture_disturb}
matreex::disturb_fun_mixt
# Giving the new disturb_fun
Picea_sp$disturb_fun <- disturb_fun_mixt
Fagus_sp$disturb_fun <- disturb_fun_mixt
```

This is an example simulation. This is important to have a biotic disturbance as well as to 

```{r jasper_sim}
disturb <- data.frame(type = "biotic", intensity = 0.2, IsSurv = FALSE, t = 500)
set.seed(42)
mixture_PiFa <- sim_deter_forest.forest(
    forest(species = list(Picea = Picea_sp, Fagus = Fagus_sp)), 
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb,
    harvest = "Uneven", targetBA = 30,
    verbose = TRUE, correction = "cut")

mixture_PiFa %>%
    filter(var %in% c("BAsp", "N"), ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

I compare the two simulations in order to check the presence of the effect. As seen above in messages, the mortality is multiplied by 0.74, which we see here with the blue line under the red one for *Picea abies*.

*Yes, the disturbance mortality is saved in the harvest variable since there is no harvest when there is a disturbance.*

```{r comparaison}
bind_rows(mixture = mixture_PiFa, basic =  disturb_PiFa, .id = "jasper") %>% 
    filter(var == "h", time == 500) %>%
    ggplot(aes(x = mesh, y = value, color = jasper)) + 
    facet_wrap(~ species, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```


# References

<div id="refs"></div>
