---
title: "Disturbance models"
package: matreex
output: 
    github_document:
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Disturbance models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib 
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(cli.progress_show_after = 600) 
# 10 minutes before showing a cli progress bar
```


This vignette illustrates how to use disturbance scenarii with `{matreex}` package. A disturbance scenario provides the timing and intensity at which different types of disturbance (storm, fire and biotic) will affect tree mortality in each size class. The tree killed by the disturbance are saved within the harvested output.

The basic `{matreex}` functions are shown in [a previous introduction vignette](matreex.html).


# Simulations input

## Define a species

The first step is the IPM integration and species object creation. This part is common with basic usage of the package, so nothing is very important here.

**Please keep in mind this computation is intensive and may take few minutes !**

```{r make_IPM_Picea}
library(matreex)
library(dplyr)
library(ggplot2)

# Load fitted model for a species
# fit_species # list of all species in dataset
data("fit_Picea_abies")

# Load associated climate
data("climate_species")
climate <- subset(climate_species, N == 2 & sp == "Picea_abies", select = -sp)
# see ?climate_species to understand the filtering of N.
climate

Picea_ipm <- make_IPM(
    species = "Picea_abies", 
    climate = climate, 
    fit = fit_Picea_abies,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:70, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
```

```{r hide warn, echo=FALSE, }
options(W_matreex_edist = FALSE)
```

Latter simulations will start at equilibrium, so we compute this before anything. `Picea_equil` is the equilibrium distribution after 2500 years.

```{r equilibrium}
Picea_sp <- species(IPM = Picea_ipm, init_pop = def_initBA(40))
Picea_for <- forest(species = list(Picea = Picea_sp))
set.seed(42)
Picea_sim <- sim_deter_forest(Picea_for, tlim = 2500,
                              equil_dist = 250, equil_time = 2500,
                              verbose = TRUE) 

Picea_equil <- filter(Picea_sim, var == "n", equil) %>% pull(value)
```


# Disturbance

## Definition of disturbance

We define a disturbance by few parameters used later in the formula. Details on the model definition and dataset used can be found in @Barrere2023.

The calibration of the equations of disturbance mortality implemented in the package is described in details in @Barrere2023. We characterize a disturbance in the simulations with two variables:

- $I$, its intensity between 0 and 1.

- $type$, the class of disturbance. This is a label in `"storm"`, `"fire"` and `"biotic"`. This is used to filter species parameters fitted, which were estimated separately for each disturbance type.

These variables are set in a data.frame object. A last column named **IsSurv** is used to inform whether the survival part of the IPM (competition-induced mortality) is needed during a disturbance. In @Barrere2023, the model does not differentiate disturbance mortality and background mortality. Therefore, we recommend to deactivate baseline mortality to avoid an overestimation of mortality. 

```{r}
ex_disturb <- data.frame(type = "storm", intensity = 0.5, IsSurv = FALSE)
```


## Impact on population

The disturbance impact on the population is calculated from parameters estimated in @Barrere2023. The disturbance mortality equation takes the size distribution, quadratic diameter of the species, intensity ($I$) and duration ($t$) of the disturbance as inputs. A set of parameters is associated to each type of disturbance and species.

$$
dqm = \sqrt{\frac{\sum_{i=1}^m size_i^2 \times value_i}{\sum_{i=1}^m value_i}} \\
logratio = log(\frac{size}{dqm}) \\
dbh.scaled = dbh.intercept + size \times dbh.slope \\
logratio.scaled = logratio.intercept + logratio \times logratio.slope \\
X_{t+1} = X_t \times (1 - plogis(a_0 + a_1 \times logratio.scaled + b \times I^{c \times dbh.scaled}))^t 
$$

The parameters are estimated with Bayesian computations. The mean of all posterior estimations are stored inside the package for each combination of species and disturbance type.

```{r data_pkg}
(Picea_coefs <- filter(matreex::disturb_coef, species == "Picea_abies"))
Picea_sp$disturb_coef <- Picea_coefs
```

Along with this set of parameters, we need to provide a disturbance function to the species we want to simulate. *The species is initiated with a default empty function that will throw warnings.* The function also require the quadratic mean diameter of the population.

```{r debug_fun, eval = FALSE, echo=FALSE}
x <- Picea_equil
species <- Picea_sp
disturb <- data.frame(type = "storm", intensity = 0.7, IsSurv = FALSE, time = 500)
qmd <-QMD(Picea_sp$IPM$mesh, x)
```

Below is an example of the difference in size distribution before and after a disturbance of type `"storm"` and intensity `0.7`.

```{r disturb_fun}
#' Disturbance function
matreex::disturb_fun
Picea_sp$disturb_fun <- disturb_fun
```


```{r disturb_ex, echo = FALSE}
ex_disturb <- data.frame(type = "storm", intensity = 0.7, IsSurv = FALSE)
qmd <-QMD(Picea_sp$IPM$mesh, Picea_equil)

plot(Picea_equil, type= "l", xlab = "size index", ylab = "n")
lines(1:(700 + as.numeric(Picea_sp$IPM$info["delay"])), 
      Picea_equil - disturb_fun(Picea_equil, Picea_sp, ex_disturb, qmd = qmd), 
      col = "red")
legend("topright", c("Distribution", "After Disturbance"),
       lty = c(1, 1), col = c(1, 2))
```

Any user can define the `disturb_fun()` function they need and provide parameters for the wanted species. Obviously, this type of modification have not been thoroughly tested and thus is more prone to error so don’t hesitate to contact `{matreex}` maintainer in this eventuality.

## Simulations

Running a simulation takes the same parameters as usual, with an added data.frame with disturbance along time. We will start from equilibrium and have a disturbance after 100 years.

```{r disturbance table}
disturb <- data.frame(type = "storm", intensity = 0.2, IsSurv = FALSE, t = 100)
```

**Disturbance size distribution is saved in harvest output. Harvest and basal mortality are canceled when a disturbance happens.**

```{r sim}
Picea_sp$init_pop <- def_init_k(Picea_equil)

disturb_picea <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp)),
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb, # providing disturbance table
    verbose = TRUE)

disturb_picea %>%
    filter(var %in% c("BAsp", "H"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

<!--
## Delay

I test delay here just to be sure. And there was a bug in get_step_IPM so I was right. 

```{r delay, eval = FALSE}
load_all()
time <- 1500
(disturb <- data.frame(type = "storm", intensity = 0.2, 
                       IsSurv = FALSE, t = 600))
forest_d <- forest(species = list(Abies = delay(Abies_alba, 5)))
set.seed(42)
memor_d <- sim_deter_forest(forest_d, tlim = time,
                                 equil_dist = 250, equil_time = time,
                                 disturbance  = disturb,
                                 harvest = "Uneven", targetBA = 30,
                                 verbose = TRUE, correction = "cut")

memor_d %>%
    filter(var %in% c("BAsp", "N"), ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(size = .4) + geom_point(size = .4) +
    NULL
```
-->

## Multispecific simulations 

Plurispecific simulations are performed by creating species object and attributing disturbance function to each species individually as previously illustrated. The disturbance dataframe is common to all species included in the simulations. We also change the harvesting model to illustrate other kind of simulations.

```{r abies_ipm}
data("fit_Fagus_sylvatica")

Fagus_ipm <- make_IPM(
    species = "Fagus_sylvatica", 
    climate = climate, 
    fit = fit_Fagus_sylvatica,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
    BA = 0:70, 
    verbose = TRUE
)
Fagus_sp <- species(
    IPM = Fagus_ipm, init_pop = def_initBA(30),
    disturb_fun = disturb_fun, 
    harvest_fun = Uneven_harv,
    disturb_coef = filter(matreex::disturb_coef, species == "Fagus_sylvatica")
)
```

```{r nsp_sim}
time <- 1500
disturb <- data.frame(type = "storm", intensity = 0.2, IsSurv = FALSE, t = 500)
Picea_sp$harvest_fun <- Uneven_harv

set.seed(42)
disturb_PiFa <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp, Fagus = Fagus_sp),
           harv_rules = c(Pmax = 0.25, dBAmin = 3, freq = 10, alpha = 1)), 
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb,
    harvest = "Uneven", targetBA = 30,
    verbose = TRUE, correction = "cut")

disturb_PiFa %>%
    filter(var %in% c("BAsp", "H"), time > 450 & time < 600, ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```


## Linear stabilizing effect of species mixture

When simulating the effect of biotic disturbances on forests mixing coniferous and broadleaf species, an additional effect can be taken into account based on the survival models developped by @Brandl2020. This mixture effect is a linear function scaled to the data used to compute disturbance coefficient present in the package (@Barrere2023).

To include this effect, the probability of disturbance mortality is multiplied by the linear function linking the share of conifer to biotic mortality.

$$
f_{mixture.effect}\left(share.conifers\right) = 0.4767 + 0.6015 \cdot share.conifers \\
h_{new}\left(share.conifers\right) = h_{disturb} \cdot f_{mixture.effect}\left(share.conifers\right)
$$

To compute the share of conifers, the type (broadleaf or conifer) of each species needs to be set as shown below. This can be set when creating the species object or edited later :

```{r type}
Picea_sp$info["type"]
Picea_sp$info["type"] <- "Coniferous"
Fagus_sp$info["type"] <- "Broadleaf"
```

An alternative disturbance function is provided by the package as `disturb_fun_mixt()`, to accept the input of the coniferous percentage computed at the plot scale before disturbing each species and to modify the mortality probability (`Pkill`) if the species is a coniferous.

```{r mixture_disturb}
matreex::disturb_fun_mixt
# Giving the new disturb_fun
Picea_sp$disturb_fun <- disturb_fun_mixt
Fagus_sp$disturb_fun <- disturb_fun_mixt
```

This is an example simulation. Note that it is imperative to apply this mixture effect with biotic disturbance types only, since this effect was only calibrated for biotic mortality. 

```{r jasper_sim}
disturb <- data.frame(type = "biotic", intensity = 0.2, IsSurv = FALSE, t = 500)
set.seed(42)
mixture_PiFa <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp, Fagus = Fagus_sp)), 
    tlim = 1000, equil_dist = 250, equil_time = 1000,
    disturbance  = disturb,
    harvest = "Uneven", targetBA = 30,
    verbose = TRUE, correction = "cut")

mixture_PiFa %>%
    filter(var %in% c("BAsp", "N"), ! equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

I compare the two simulations in order to check the presence of the effect. As seen above in messages, the mortality is multiplied by 0.74, which we see here with the blue line under the red one for *Picea abies*.

*Yes, the disturbance mortality is saved in the harvest variable since there is no harvest when there is a disturbance.*

```{r comparaison}
bind_rows(mixture = mixture_PiFa, basic =  disturb_PiFa, .id = "jasper") %>% 
    filter(var == "h", time == 500) %>%
    ggplot(aes(x = mesh, y = value, color = jasper)) + 
    facet_wrap(~ species, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```


# References

<div id="refs"></div>
