---
title: "Harvesting models"
package: matreex
output: 
    github_document:
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Harvesting models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib 
link-citations: true
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(cli.progress_show_after = 600) 
# 10 minutes before showing a cli progress bar

library(tidyr)
```

This vignette illustrates how to use harvesting scenarii with `{matreex}` package. An harvesting scenario provides the rules and the periodicity to trigger harvesting of trees in each size classes. The harvested trees are saved and exported for each simulation time step.

The basic `{matreex}` functions are shown in [a previous introduction vignette](matreex.html).

# Simulations input

## Define a species

The first step is the IPM integration. This part is common with basic usage of the package, so nothing is very important here.

**Please keep in mind this computation is intensive and may take few minutes !**

```{r make_IPM_Picea}
library(matreex)
library(dplyr)
library(ggplot2)

# Load fitted model for a species
# fit_species # list of all species in dataset
data("fit_Picea_abies")

# Load associated climate
data("climate_species")
climate <- subset(climate_species, N == 2 & sp == "Picea_abies", select = -sp)
# see ?climate_species to understand the filtering of N.
climate

Picea_ipm <- make_IPM(
    species = "Picea_abies", 
    climate = climate, 
    fit = fit_Picea_abies,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:70, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
```

Harvesting scenario are defined at different levels: at the species level with the species harvesting function and its associated parameters, at the forest level with a set of parameters, and at the simulation level with the target of the harvesting scenario. Depending on the scenario used, not all the parameters are used. The table below summarise the required parameters for each scenario. Each scenario have an example in the following document.

| Harvesting scenario | Species harvesting function | Species parameters | Forest parameters | Simulation parameters |
| -- | -- | -- | -- | -- |
| default | `def_harv()` <br> or custom by user. | $\emptyset$ | `harv_rules["freq"]` | $\emptyset$ |
| Uneven | `Uneven_harv()` | `harv_lim` | `harv_rules` | `targetBA` |
| Even  <br> *in active dev* | `Even_harv()` | `rdi_coef` | `harv_rules` <!--`FinalHarvT`--> | `targetRDI` <br> `targetKg` |

Short example : When using an Uneven scenario, each species needs to use the `Uneven_harv()` function with `harv_lim` parameters. The forest will need to have `harv_rules` parameters. The `sim_deter_forest()` require `targetBA` argument. If `targetRDI` is defined, it will not be used.

# Default scenario

## Presentation

The default scenario is based on @kunstler2021. This mean there is a constant harvest rate triggered each year. This harvest rate is uniform over the size distribution. This is the scenario used by default.

This rate is coded in `def_harv()` function as shown below and the frequency (each year) is given in `harv_rules["freq"]`. Note that the harvesting is applied to all size class except the age-classes used in delay (with `* (ct > 0)`). 


```{r def_Picea}
def_harv
Picea_sp <- species(IPM = Picea_ipm, init_pop = def_initBA(30))
Picea_for <- forest(species = list(Picea = Picea_sp), 
                    harv_rules = c(Pmax = 0.25, dBAmin = 3, 
                                   freq = 1, alpha = 1))
```

With this scenario, the simulation can be launched without any additional parameter.

```{r sp1sim}
set.seed(42) # The seed is here to replicate the random population initialisation.
Picea_sim <- sim_deter_forest(
    Picea_for,
    tlim = 200,
    equil_time = 200, equil_dist = 10, equil_diff = 1,
    harvest = "default", # this is the default value but we write it.
    SurfEch = 0.03,
    verbose = TRUE
)
```

Once the simulation is done, we can extract the basal area and the number of individual at each step of the simulation, but also a new variable $H$. This is the sum of density of harvested trees at each time step. The density of harvested trees is also exported for each size mesh $h_i$.

In this case, $H$ is correlated with $N$ since it's a constant percentage not linked with a size distribution. The first step being the initialization step, it's normal to have no harvest.

```{r sp1plot}
Picea_sim  %>%
    dplyr::filter(var %in% c("BAsp", "N", "H"), ! equil) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .2) + geom_point(size = 0.4) 
```

## Modulation

This section will just illustrate variation of the default scenario. First we modify the frequency of the harvest. When the harvest is not triggered, the value returned is 0.

```{r def_freq}
set.seed(42) # The seed is here for initial population random functions.
Picea_sim_f20 <- sim_deter_forest(
    forest(species = list(Picea = Picea_sp), 
                      harv_rules = c(Pmax = 0.25, dBAmin = 3, 
                                     freq = 20, alpha = 1)),
    tlim = 50,
    equil_time = 50, equil_dist = 10, equil_diff = 1,
    harvest = "default", 
    SurfEch = 0.03,
    verbose = TRUE
)
Picea_sim_f20  %>%
    dplyr::filter(var %in% c("BAsp", "N", "H"), ! equil) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .2) + geom_point(size = 0.4) 
```

A more advanced modification of the harvest function is ilustratted below. Obviously, this type of modification have not been thoroughly tested and thus is more prone to error so don't hesitate to contact `{matreex}` maintainer in case of trouble. For example, we add a function where we multiply a constant rate with the mesh, meaning that the larger the tree get, the higher the harvesting rate of its size class.

```{r edit_def_harv}
Picea_harv <- Picea_sp
Picea_harv$harvest_fun <- function(x, species, ...){

    dots <- list(...)
    ct <- dots$ct

    rate <- 6e-4 * (ct > 0) * species$IPM$mesh 
    return(x * rate)
}

set.seed(42) # The seed is here for initial population random functions.
Picea_sim_f20 <- sim_deter_forest(
    forest(species = list(Picea = Picea_harv), 
                      harv_rules = c(Pmax = 0.25, dBAmin = 3, 
                                     freq = 20, alpha = 1)),
    tlim = 250,
    equil_time = 250, equil_dist = 10, equil_diff = 1,
    harvest = "default",
    SurfEch = 0.03,
    verbose = TRUE
)
Picea_sim_f20  %>%
    dplyr::filter(var %in% c("BAsp", "N", "H"), ! equil) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .2) + geom_point(size = 0.4) 
```

# Uneven scenario

## Theory

Uneven-aged harvest scenario consist in harvesting trees in all size classes with the objective to reach a stable size structure with continuous replacing  of large mature trees. This scenario depends on the basal area of the stand and the size distribution of the tree. This should lead to stands with uneven size distribution.

### Monospecific case

#### Harvest proportion

We note $P_{cut}$ the global harvest proportion which will determine the amount of basal area to be cut and harvested.

$$
P_{cut} = \left\lbrace
 \begin{array}{ll}
    0 & if (BA_{stand} - BA_{target}) < \Delta BA_{min}\\
    min(\frac{BA_{stand}-BA_{target}}{BA_{stand}}, P_{max}) & if (BA_{stand} - BA_{target}) \geq \Delta BA_{min}
 \end{array}
 \right\rbrace
$$

For example, some numerical values of the parameters can be $\Delta BA_{min} = 3 m^2ha^{-1}$, $BA_{target} = 20, 25$ or $30 m^2ha^{-1}$ (depending on species) and $P_{max} = 0.25$.

Note that stand basal area $BA_{stand}$ is computed only considering trees with a dbh above $d_{th}$ (see below).

#### Harvest curve

Each tree harvest probability only depends on its diameter ($d$).
There is a minimum diameter for harvest ($d_{th}$), harvest probability then increases with diameter until $d_{ha}$ after which harvest probability is constant.

We therefore considered the harvesting function (which associates a dbh to an harvesting probability)

$$
h(d) = \left\lbrace
 \begin{array}{ll}
    0 & \text{if } d < d_{th} \\
    h_{max} (\frac{d - d_{th}}{d_{ha} - d_{th}})^{k} & \text{if } d_{th}\leq d < d_{ha} \\
    h_{max} & \text{if } d \geq d_{ha} 
 \end{array}
 \right\rbrace
$$

The maximum harvesting for large tree $h_{max}$ can be tuned so that the probability for a large tree to be harvested approaches  1 after several harvesting operations:
$h_{max} =  1 - \sqrt[n]{1-p}$ with $n$ the number of harvesting operations. Parameter $k$ defines how quickly the harvesting rate increase from $d_{th}$  to $d_{ha}$.

```{r setplot_val, echo = FALSE}
library(latex2exp)
dth <- 17.5
dha <- 57.5
hmax <- 0.8
k <- 2
```

```{r, harvest_curve_plot, echo = FALSE, warning = FALSE, fig.cap = sprintf("Harvest curve example, $d_{th} = %.1fcm$, $d_{ha}=%.1fcm$, $h_{max}=%.1f$, $k=%.0f$.", dth, dha, hmax, k)}
d <- seq(0, 100, length.out=100)
h <- function(d){
   harvecurve <- hmax * ((d - dth)/(dha-dth))^k
   harvecurve[d<dth] <- 0
   harvecurve[d>dha] <- hmax
   return(harvecurve)
}
df <- data.frame(dbh=d, h=h(d))
ggplot(df, aes(x=dbh, y=h)) + geom_line(col='red') + theme_bw() +
	geom_label(x=dth, y=0.05, label=TeX("$d_{th}$")) +
	geom_label(x=dha-5, y=0.05, label=TeX("$d_{ha}$")) +
	geom_label(x=dha-5, y=hmax, label=TeX("$h_{max}$")) +
	geom_vline(aes(xintercept=dha), linetype=2)
```

#### Harvesting algorithm

The algorithm only target tree contributing to $BA_{stand}$, that is the trees above $d_{th}$

Given $\phi(x)$ the density function of diameters, the basal area harvested is

$$ 
BA_{harv} = BA_{th} + BA_{ha}
$$
with 
$$ 
BA_{ha} =  h_{max} \pi/4\int_{d_{ha}}^{d_{max}}x^2 \phi(x)dx
$$

$\bullet$ If $BA_{ha} >= P_{cut} \times BA_{stand}$, there is enough large trees (diameter above $d_{ha}$) so that the harvest will only concern large trees and $BA_{th} = 0$.

We then find $d_t$ (with $d_{ha} < d_t < d_{max}$) such that:
$$  BA_{ha} = h_{max} \pi/4 \int_{d_{t}}^{d_{max}}x^2 \phi(x)dx = P_{cut} \times BA_{stand}$$
Thus, the `{matreex}` package will cut the larger trees until $P_{cut} \times BA_{ha} - targetBA <= 0$.


$\bullet$ If $BA_{ha} < P_{cut} \times BA_{stand}$, we first harvest $BA_{ha}$ and then compute $k$ such as: $BA_{th} = P_{cut} \times BA_{stand} - BA_{ha}$. Note that

### Multispecific case

<!-- #### Abundance-based preference -->

As in the monospecific case, we define the global harvest rate $P_{cut} = \frac{BA_{harv}}{BA_{stand}}$.

Here, $BA_{stand}$ is divided between $s$ species: 
$BA_{stand} = \sum_{i=1}^{s} BA_{stand, i}$ and
$BA_{harv} = \sum_{i=1}^{s} BA_{harv, i}$

We note $p_i = \frac{BA_{stand, i}}{BA_{stand}}$ and 
$P_{cut, i} = \frac{BA_{stand, i} - BA_{harv, i}}{BA_{stand, i}} = f(p_i) * P_{cut}$

We suppose that harvesting rate increases with abundance (we harvest preferentially trees with the highest proportion), which means $f$ is an increasing function.

By definition,

$$
\begin{array}{ll}
BA_{harv} & = \sum_{i=1}^{s} BA_{harv, i} = \sum_{i=1}^{s} BA_{stand, i} * (1 - P_{cut, i}) \\
          & = BA_{stand} \sum_{i=1}^{s} p_i * (1 - f(P_i) P_{cut}) \\
	  & = BA_{stand} (1 - P_{cut}) \\
\end{array}
$$

So that we have the constraint on $f$: 
$\sum_{i=1}^{s} p_i (1-f(p_i)P_{cut}) = 1 - P_{cut} \sum_{i=1}^{s} p_i f(p_i) = 1-P_{cut}$

which is equivalent to
$\sum_{i=1}^{s} p_i f(p_i) = 1$

The case $f(p_i) = 1$ works, which leads to $P_{cut,i} = P_{cut}$. In that case the harvest rate is the same for every species $i$.

More broadly, we can use the function $$f(p_i) = \frac{p_i^{\alpha - 1}}{\sum_{i=1}^{s} p_i ^{\alpha}}$$, $ \forall \alpha > 0$

For $\alpha = 2$, we for example have 

$$f(p_i) =\frac{p_i}{\sum_{i=1}^{s} p_i ^2} $$
To sum up, if$\alpha = 1$, an abundant species will be more harvested, in effort to reach an equi-abundance of the species. For $\alpha < 1$, the abundant species will be less harvested. Finally, for $\alpha > 1$ the most abundant species will be harvested more, in a greater proportion than 1.

<!--
#### Favoured species

In some case, we may want to favour some species compared to others.
We note $Q$ the $q$ species we want to favour, and $P_Q=\sum_{i=1}^{q} p_i$
We first compute the harvest rate $H_Q$ and $H_{1-Q}$ for respectively the favoured/other species.

If $P_Q \geq 0.5$, we take $H_Q = H_{1-Q} = H$ (the species to be favoured are already dominant).


If $P_Q < 0.5$, we compute $H_Q=f(P_Q) H$ and $H_{1-Q}=f(1-P_Q) H$ with $\alpha > 1$.
By definition, we will get $H_Q \leq H$.

We then apply for each species $i$ the harvest rate $H_i = H_Q$ or $H_i=H_{1-Q}$ depending on which group it belongs to.
-->


## Examples

All the parameters described above are input either in the `species()`, `forest()` or `sim_deter_forest()` functions. Additional parameter `dBAmin` is the difference between `BA` and `targetBA` under which an harvesting will not be triggered. 

```{r Uneven_Picea}
Picea_Uneven <- species(IPM = Picea_ipm, init_pop = def_initBA(30), 
                        harvest_fun = Uneven_harv,
                        harv_lim = c(dth = 175, dha = 575, hmax = 1))
Picea_for_Uneven <- forest(species = list(Picea = Picea_Uneven), 
                      harv_rules = c(Pmax = 0.25, dBAmin = 3, 
                                     freq = 5, alpha = 1))
```



```{r Uneven_sim}
set.seed(42) # The seed is here for initial population random functions.
Picea_sim_f20 <- sim_deter_forest(
    Picea_for_Uneven,
    tlim = 260,
    equil_time = 260, equil_dist = 10, equil_diff = 1,
    harvest = "Uneven", targetBA = 20, # We change the harvest and set targetBA.
    SurfEch = 0.03,
    verbose = TRUE
)
Picea_sim_f20  %>%
    dplyr::filter(var %in% c("BAsp", "N", "H"), ! equil) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .2) + geom_point(size = 0.4) 
```

We notice that the basal area obtained by the simulation is higher than the targeted one. This can be explained by the fact that the cutting calculation is done on $BA_{stand}$, which does not take into account individuals smaller than $d_{th}$.

# Even scenario

<!-- $$ -->
<!-- dg = \sqrt{\frac{\sum_{i = 0}^n d_i^2 x_i}{ \sum_{i = 0}^n x_i }} \\ -->
<!-- dg_{cut} = \sqrt{\frac{\sum_{i = 0}^n d_i^2 x_i Pc_i }{ \sum_{i = 0}^n x_i Pc_i }} -->
<!-- $$ -->

<!-- mesh in cm. -->
<!-- $$ -->
<!-- RDI = \frac{ \sum_{i = 0}^n x_i }{ e^{ RDI_{int} + RDI_{slope} \times \frac{ \log( \frac{ \sum_{i = 0}^n mesh_i^2  x_i }{ \sum_{i = 0}^n x_i} ) }{2}   } } -->
<!-- $$ -->

The objective of even harvesting is to grow trees with the same age until they reach a self-thinning boundary and cut all of them at this stage. This self-thinning boundary is given for each species following :

$$N_{max} = e^{intercept+slope \cdot log(Dg)}$$
The species parameters are given in @Aussenac2021 and inside the package with `rdi_coef` table. $Dg$ is the mean quadratic diameter of trees.

```{r rdi_coef}
data(rdi_coef)
rdi_coef <- drop(as.matrix(
    rdi_coef[rdi_coef$species == "Picea_abies",c("intercept", "slope")]
))
rdi_coef
```

From this, we can compute the density index $DI = N / N_{max}$ with $N$ the number of stems. 

During the growth process, trees below the mean quadratic diameter $Dg$ are cut. The magnitude of this cut is given with $Kg$, the ratio between the mean quadratic diameter of killed trees $Dg_d$ and the mean quadratic diameter $Dg$ :

$$Kg = Dg^2_d / Dg^2$$

The algorithm present in `Even_harv` is triggered if $DI > targetRDI$ and
optimise two parameters $h_{max}$ and $k$ for a cut probability curve to reach $targetKg$ and $targetRDI$.

$$Pcut_i = hmax * i^{-k}$$

```{r, even_hcurve_plot, echo = FALSE, warning = FALSE, fig.cap = "Harvest curve example with various combination of hmax and k"}
h_fun <- function(hmax, k,n = 100){
    d <- seq(0, n, length.out=n)
   harvecurve <- hmax * (d)^(-k)
   return(harvecurve)
}

n = 100

expand.grid(hmax = c(.5, .8), k = c(0.1, .5)) %>%
    rowwise() %>%
    mutate(pcut = list(h_fun(hmax, k, n)), dbh = list(1:n)) %>%
    tidyr::unnest(c(pcut, dbh)) %>%
    mutate(hmax = as.factor(hmax), k = as.factor(k)) %>%
    ggplot( aes(x=dbh, y=pcut, col = hmax,  linetype = k)) + 
    geom_line() +
    theme_bw() +
    NULL
```


The last parameter for an even harvested simulation is the final cut time. The stand will be harvested with the previous rules at a given frequency but at a certain point, the manager will harvest all trees and plant new ones. This time is dependant on the species growth and differents targets. The value is named `final_harv` in `sim_deter_forest()` function.

**Because the rdi coefficient depends on a species, it's not possible to use multiple species forest with {matreex} package. Please contact authors if you need to use mutlispecific even harvesting.**

<!-- Mortality is triggered when stand density exceeds the self-thinning boundary, i.e. when $DI$  is greater than or equal to 1. In that case, $DI$ is reduced to 1 using the $Kg$ parameter, with -->

<!-- To reduce $DI$ to 1, the number of trees in the stand is reduced while the mean quadratic diameter is increased, which amounts to preferentially killing small trees. In mixed stands, the number of trees killed for each species is defined according to their proportion in the stand before mortality process. -->


## Examples

The rdi parameters are input in the `species()` function. Calling `forest()` function is not different, only the frequency of harvest will be used. The different targets `targetKg`, `targetRDI` and `final_harv` values are set when launching a simulation.

```{r Even_Picea}
Picea_Even <- species(
    IPM = Picea_ipm, init_pop = def_init_even,
    harvest_fun = Even_harv, rdi_coef = rdi_coef,
    harv_lim = c(dth = 175, dha = 575, hmax = 1)
)

Picea_for_Even <- forest(species = list(Picea = Picea_Even),
                      harv_rules = c(Pmax = 0.25, dBAmin = 3,
                                     freq = 5, alpha = 1))
```


```{r Even_sim}
set.seed(42) # The seed is here for initial population random functions.
Picea_sim_f20 <- sim_deter_forest(
    Picea_for_Even,
    tlim = 100,
    equil_time = 100, equil_dist = 10, equil_diff = 1,
    harvest = "Even", targetRDI = 0.9, targetKg = 0.6,
    final_harv = 80,
    SurfEch = 0.03,
    verbose = TRUE
)
Picea_sim_f20  %>%
    dplyr::filter(var %in% c("BAsp", "N", "H"), ! equil) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .2) + geom_point(size = 0.4)
```


# References

<div id="refs"></div>
