---
title: "Individual based model version"
package: matreex
output: 
    github_document:
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Individual based model version}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib 
link-citations: true
---

```{r, include = FALSE}
# load_all()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(cli.progress_show_after = 600) 
# 10 minutes before showing a cli progress bar
```

This vignette illustrates the difference between functions used to run simulations with IPM and IBM in the `{matreex}` package. The package can model the dynamic of monospecific or plurispecific tree communities. IBM is simpler than IPM in its use of the different modules for harvesting and disturbances. 

The basic `{matreex}` functions are shown in [a previous introduction vignette](matreex.html).

# Rapid description of the Individual Based Model

Details on the fitting and the integration of IPM model can be found in @ellner2016.
Briefly, an IPM predicts the size distribution, $n(z', t+1)$, of a population at time t+1 from its size distribution at time t, $n(z, t)$, with $z$ the size at t and $z'$ the size at $t + 1$,
based on the following equation (@ellner2016): 

$$
n(z', t+1) = \int_{L}^{U} K(z', z) n(z, t) dz 
$$

with $L$ and $U$ being, respectively, the lower and upper observed sizes for integration of the kernel $K$.

The kernel $K(z' ,z)$ can be split into the survival and growth kernel $P(z' ,z)$ and the fecundity kernel $F(z', z)$, as follows : $K(z' ,z) = P(z' ,z) + F(z' ,z)$ .

The IBM use the same functions as the kernels but apply them on numerical vectors of individual sizes instead of a continuous size distribution $n$. Details on the numerical integration are given below and in @kunstler2021 and @guyennon2023 . 
<!--The figure 1 presents the structure and approach of the IPM. -->
Note, that this package do not cover the statistical fitting of the vital rates functions.


# Simulations input

## Define a species

The first step is the IPM integration. This part is common with basic usage of the package, so nothing is very important here. 

**Please keep in mind this computation is intensive and may take few minutes !**
If you only use the IBM function to simulate later, the IPM integration can be done for a single basal area `BA = 0`.

```{r make_IPM_Picea}
# library(matreex)
devtools::load_all()
library(dplyr)
library(ggplot2)

# Load fitted model for a species
# fit_species # list of all species in dataset
data("fit_Picea_abies")

# Load associated climate
data("climate_species")
climate <- subset(climate_species, N == 2 & sp == "Picea_abies", select = -sp)
# see ?climate_species to understand the filtering of N.
climate

Picea_ipm <- make_IPM(
    species = "Picea_abies", 
    climate = climate, 
    fit = fit_Picea_abies,
    clim_lab = "optimum clim", 
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Picea_abies) * 1.1),
    BA = 0:80, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
```

```{r hide warn, echo=FALSE, }
options(W_matreex_edist = FALSE)
```

The IPM is used in order to have a common base with other functions. However, the IBM simulation does not allow to use some modules from the package :

* `init_pop` : When initiating the size distribution, the continuous distribution is draw for a basal area and transformed into an individual distribution via a poisson distribution inside the size classes (`def_init_pop()`). It's possible to rewrite this function by replacing the first line with other initiation functions of the package.  `def_init_k()` is discussed later on.

```{r}
def_init_pop
```


* `harvest_fun` : The harvest is only possible with a constant rate given as entry of the simulation function.

* `disturb_fun` : No disturbance is possible.

For this example, we will just modify the initial size distribution to start at a basal area of 30 $m^2/ha$.

```{r species}
Picea_sp <- species(IPM = Picea_ipm, mat_size = "sp")
```

## Define a forest

Once each species objects are built, we can assemble them in a `forest` object.

```{r forest}
Picea_for <- forest(species = list(Picea = Picea_sp))
```

# Running simulations

A simulation runs as before but ends at `tlim`, because no equilibrium is reachable by definition. The simulated surface `SurfEch` is an important parameter because it will drive the drift toward extinction. The default value is $300m^2$ since the models were fitted on this size of plot.

For this example, we will just modify the initial size distribution to start at a basal area of 30 $m^2/ha$.

```{r sp1sim}
Picea_for$species$Picea_abies$init_pop <- def_initBA(30)
set.seed(42) # The seed is here for initial population random functions.
Picea_sim <- sim_deter_forest(
    Picea_for, 
    tlim = 300, 
    equil_time = 300, equil_dist = 50, equil_diff = 1,
    SurfEch = 0.03,
    verbose = TRUE
) 
Picea_for$species$Picea_abies$init_pop <- def_initBA_pop(30)
set.seed(42)
Picea_indiv <- sim_indiv_forest(Picea_for, tlim = 300,  SurfEch = 0.03,
                                harvest = 0.006, verbose = TRUE)
```

In the code above, we simulate for 200 years (`tlim`) years

The output of a simulation is a data.frame in long format (according to tidyverse style). This is very helpful to filter the output and plot it with `{ggplot2}`. Variables exported are the basal area per species `BAsp`, and `N` and `H` the total number of alive and dead individuals in the forest (per hectare). Individual information are not saved.

Note that we need to scale down the deterministic simulation in order to compare both simulations.

```{r sp1plot}
All_sims <- dplyr::bind_rows(deter = Picea_sim, indiv = Picea_indiv, .id = "type")
All_sims %>%
    dplyr::filter(var %in% c("BAsp", "N")) %>%
    # modify output for SurfEch and not per ha in the deterministic sim.
    dplyr::mutate(value = ifelse(type == "deter" & var == "N", 
                                 value * 0.03, value)) %>%
    ggplot(aes(x = time, y = value, color = type)) +
    geom_line(linewidth = .4) + 
    facet_wrap(~ var, scales = "free_y")
```

# Customizing the simulations

## Multiple species

Multi-specific simulations are performed like the simulations previously illustrated. The only difference is in the construction of the forest object. This explain why the `species` argument for `forest()` function require a list for input. 

We need to modelise a second species. Be careful to select the same climate as the first species.

```{r second_species}
data("fit_Abies_alba")

Abies_ipm <- make_IPM(
    species = "Abies_alba", 
    climate = climate, # this variable is defined at the top of the doc.
    fit = fit_Abies_alba,
    clim_lab = "optimum clim",
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Abies_alba) * 1.1),
    BA = 0, # Default values are 0:200, smaller values speed up this vignette.
    verbose = TRUE
)
Abies_sp <- species(IPM = Abies_ipm, init_pop = def_initBA_pop(35), 
                    mat_size = "sp")
```

```{r nsp_forest}
# We edit back the init_fun for Picea
Picea_sp$init_pop <- def_initBA_pop(15)
Picea_Abies_for <- forest(species = list(Picea = Picea_sp, Abies = Abies_sp))

set.seed(8)
Picea_Abies_sim <- sim_indiv_forest(
    Picea_Abies_for, 
    tlim = 500, 
    SurfEch = 0.03,
    harvest = 0.006,
    verbose = TRUE
)

Picea_Abies_sim  %>%
    dplyr::filter(var %in% c("BAsp", "N"), !equil) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    geom_line(linewidth = .4) +
    facet_wrap(~ var, scales = "free_y") +
    stat_summary(fun = "sum",  aes(col="Total"),
                 geom ='line', linetype = "dashed", linewidth = .3)
```

## Climate variation

This is only to test if climate variation is possible. Answer is yes but, need to hack the IPM class before simulation.

```{r climate}
# future function for the package matreex. 
# I may need some help on what's needed or usefull

library(checkmate)
#' @param x a named values for a climatic variable. 
#' @param inv_null If TRUE, the inverse variable is \code{varb = 1/(var+1)}, in
#' case of var can take 0 for value.
#' Example : \code{x = c(sgdd = 2000)}
value2sqb <- function(x, inv_null = FALSE){
    
    nms <- names(x)
    res <- numeric(3)
    names(res) <- c(nms, paste0(nms, c("2", "b")))
    res[nms] <- x
    res[paste0(nms, "2")] <- x^2
    res[paste0(nms, "b")] <- 1/(x + inv_null)
    
    res
}

#' @param climate Named vector of climatic variables.
#' @param inv_null If TRUE, the inverse variable is \code{varb = 1/(var+1)}, in
#' case of var can take 0 for value.
#' Example : \code {climate = c(sgdd = 2000, wai = 0.16) ; in_null = c(sgdd = FALSE, wai = TRUE)}
expand_clim <- function(climate, inv_null){
    
    assertNumeric(climate, any.missing = FALSE)
    assertLogical(inv_null, any.missing = FALSE)
    
    nms <- names(climate)
    res <- vector("list", length(climate))
    for(x in seq_along(nms)){
        res[[x]] <- value2sqb(climate[nms[x]], inv_null[nms[x]])
    }
    res <- unlist(res)
    
    return(res)
}
# May be usefull to create the same function but to work on data.frame 
# or matrices likes below code
tlim <- 300

wai <- climate$wai + runif(tlim, -0.5, 0.5)
wai <- pmax(wai, 0) ; wai <- pmin(wai, 1) # range the values after runif
sim_clim <- as.matrix(data.frame(
    sgdd = climate$sgdd + runif(tlim, -500, 500),
    wai = wai, 
    t = 1:tlim
))
# We need to specify if variables can be set 0 for variable compution of
# varb = 1 / var. If true varb = 1 / (var + 1)
inv_null <- c(sgdd = FALSE, wai = TRUE)

exp_clim <- vector("list", tlim)
for(i in 1:tlim){
    exp_clim[[i]] <- c(expand_clim(sim_clim[i, -3], inv_null), t = i)
}
exp_clim <- do.call("rbind", exp_clim)
exp_clim[c(1:3, 297:300),]


set.seed(42)
class(Picea_for$species[[1]]$IPM) <- "fake_ipm"
Picea_indiv_clim <- sim_indiv_forest(Picea_for, tlim = tlim,  SurfEch = 0.03,
                                climate = exp_clim,
                                harvest = 0.006, verbose = TRUE)

All_sims <- dplyr::bind_rows(indiv = Picea_indiv, clim = Picea_indiv_clim, 
                             .id = "type")
All_sims %>%
    dplyr::filter(var %in% c("BAsp", "N")) %>%
    # modify output for SurfEch and not per ha in the deterministic sim.
    dplyr::mutate(value = ifelse(type == "deter" & var == "N", 
                                 value * 0.03, value)) %>%
    ggplot(aes(x = time, y = value, color = type)) +
    geom_line(linewidth = .4) + 
    facet_wrap(~ var, scales = "free_y")
```


A function to extract the time to extinction will be added to the package. It require the simulation and output the time when all populations are equal to 0. If there is no extinction, this function return NA.

```{r}
class(Picea_for$species[[1]]$IPM) <- "ipm"
set.seed(9)
# extinct on 9
Picea_indiv <- sim_indiv_forest(Picea_for, tlim = 300,  SurfEch = 0.03,
                                harvest = 0.006, verbose = TRUE)

max(Picea_indiv$time)
Picea_indiv$time[Picea_indiv$equil]

extinct(Picea_indiv)

Picea_indiv %>%
    dplyr::filter(var == "N") %>%
    ggplot(aes(x = time, y = value)) +
    geom_line(linewidth = .4) 
```

## Time to extinction

The main goal of IBM in the present package is to compute extinction time for tree species. Previous work was done in Arnaud paper, based on @Grimm2004 intrinsic mean time to extinction.

For these simulations, the stand needs to be initiated with the same population each time. To replace `def_init_k()` function, `def_init_kpop` is defined and require a population vector with a size value for each individual.

* $size \in [1; inf[$ : diameter in $mm$ for a single tree

* $size \in ]-lag; -1]$ : coded value for individual in the lag. When a new individual is recruited, it will start at the maximum lag value ($-lag$) and grow with $s_{t+1} = s_t + 1$. When $s_t = 1$, $s_{t+1} = 90$.

* $size = 0$ : coded value for a dead individual. These individuals are removed from memory during a simulation.

```{r}
Picea_for$species[[1]]$init_pop <- def_init_kpop(x = c(-12, -5, 100, 300))
```


This is an example where I compute the intrinsic mean time to extinction. We can see that the starting population may not be established.

```{r, eval=FALSE}
# test for mean time 
n <- 50
sims <- vector("list", n)
for(i in seq_along(sims)){
    print(i)
    sims[[i]] <- sim_indiv_forest(Picea_for, tlim = 1000,  SurfEch = 0.01,
                                harvest = 0.006, verbose = FALSE)
}
ext <- purrr::map_dbl(sims, extinct)
sum(is.na(ext)) / n * 100 # number of viable sims
sum(ext[!is.na(ext)] == 1) / n * 100 # number of sim extinct time 1.

# plot of all simulations
all_sims <- bind_rows(sims, .id = "sim") 
all_sims %>%
    filter(!equil, var == 'N') %>%
    mutate(ext = is.na(rep(ext, each = 1000))) %>%
    ggplot(aes(x = time, y = value, group = sim, color = ext)) +
    geom_line(alpha = .3) 
    
e <- ext[!is.na(ext)]
hist(e)
em <- data.frame(date = e) %>% 
    group_by(date) %>% count(name = "nb") %>% ungroup() %>% 
    mutate(p_t = cumsum(nb) /sum(n), ln = -log(1 - p_t))
ggplot(em, aes(x = date)) + 
    geom_line(aes(y = ln)) 

lin <- lm(ln ~ date, em)
(Tm <- 1/lin$coefficients["date"])
```


# References

<div id="refs"></div>
