---
title: "Climate variation"
package: matreex
output: 
    github_document:
    rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Climate variation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(cli.progress_show_after = 600) 
# 10 minutes before showing a cli progress bar

library(tidyr)
```

This vignette illustrates how to use climate variation scenarii with `{matreex}` package. A climate variation scenario provides simulation context where climatic variable (*sgdd* and *wai*) may fluctuate. This type of scenario require a different integration methods that is explained below.

The basic `{matreex}` functions are shown in [a previous introduction vignette](matreex.html).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


# $\mu_G$ integration principle

When we integrate an IPM, we compute the distribution of $z_{t+1}$ for each value of $z_t$. This distribution is based on a normal kernel as : 

$$f(log(\frac{z_{t+y} - z_t}{y})) \sim \mathcal{N}(\mu,\sigma^2)$$

Here, $f(log(\frac{z_{t+y} - z_t}{y}))$ is given for each climate situation, and it is the computation of the distribution that is the longest part of the integration. This limit give us a very long time if we need to integrate a matrix per year and basal area, in case of climate variation.

However we observe that there are two dimension here : a distribution of $\mu_G$ for each climate and a distribution of $z_{t+1}$ for each $\mu_G$ value. On the schematic illustration below, we can observe that depending on the climate $\alpha$ and $\beta$, different $\mu$ values are given on the $z_t$ axis, but the $z_{t+1}$ will be the same for a given $\mu$ value.

```{r mu_distribution, engine='tikz', fig.ext = 'png', engine.opts = list(template = "tex/tikz.tex"), echo = FALSE, eval=TRUE, fig.cap="Schematic illustration of difference between $\\mu_G$ distributions and distribution of $z_{t+1}$ for a given $\\mu value.$"}
\begin{tikzpicture}%[scale = 0.5, every node/.style={scale=0.5}]
    
    \draw(2.5,0) node[above] {Climate $\alpha$};
    \fill[red!50] (0, -2) -- (5,-2) -- (5, -1) -- (0,-1);
    \draw (0,-2) grid[step=.5] (5,-1);
    \draw (2.5,-1) node[above]{$z_t$} ;
    \draw (-0.25,-1.5) node[left]{$\delta_{z}$} ;
    \draw(0,-1) node[above] {$L$};
    \draw(5,-1) node[above] {$U$};
    \draw[->] (5.25,-1) -- (5.25,-2);
    \draw(5.25,-1) node[right] {0};
    \draw(5.25,-2) node[right] {$N_{int}$};
    
    \draw (0,-3) -- (5,-3);
    \draw [->] (0,-3) -- (0,-5.5);
    
    \begin{scope}[xshift=0cm,yshift=-7.3cm]
        \draw[line width=1pt, fill=teal, color=teal] (0.75,4.25) .. controls (0.87,4.12)  and (1,3.9) .. (1,3.8)
        .. controls (1,3.75) and (0.8, 3.62) .. (0.8,3) -- (0.75, 1.8);
    \end{scope}
    \begin{scope}[xshift=2cm,yshift=-7.3cm]
        \draw[line width=1pt, fill=cyan, color=cyan] (0.75,4.25) .. controls (0.87,4.12)  and (1,3.9) .. (1,3.8)
        .. controls (1,3.75) and (0.8, 3.62) .. (0.8,3) -- (0.75, 1.8);
    \end{scope}
    
    \foreach \x in {.25,.75,...,4.75} {%
    \draw ($(\x,-3) + (0,-2pt)$) -- ($(\x,-3) + (0,2pt)$)
    node [above] {};
    }
    \draw(0.75,-3) node[above] {$x$};
    \draw(2.75,-3) node[above] {$y$};
    \draw(5,-3) node[right] {$\mu_G$};
    \draw(0,-5.5) node[left] {$\delta_z$};
    
    \begin{scope}[xshift=7cm,yshift=0cm]
        \draw(2.5,0) node[above] {Climate $\beta$};
        \fill[red!50] (0, -2) -- (5,-2) -- (5, -1) -- (0,-1);
        \draw (0,-2) grid[step=.5] (5,-1);
        \draw (2.5,-1) node[above]{$z_t$} ;
        \draw (-0.25,-1.5) node[left]{$\delta_{z}$} ;
        \draw(0,-1) node[above] {$L$};
        \draw(5,-1) node[above] {$U$};
        \draw[->] (5.25,-1) -- (5.25,-2);
        \draw(5.25,-1) node[right] {0};
        \draw(5.25,-2) node[right] {$N_{int}$};
        
        \draw (0,-3) -- (5,-3);
        \draw [->] (0,-3) -- (0,-5.5);
        
        \begin{scope}[xshift=1cm,yshift=-7.3cm]
            \draw[line width=1pt, fill=teal, color=teal] (0.75,4.25) .. controls (0.87,4.12)  and (1,3.9) .. (1,3.8)
            .. controls (1,3.75) and (0.8, 3.62) .. (0.8,3) -- (0.75, 1.8);
        \end{scope}
        \begin{scope}[xshift=2cm,yshift=-7.3cm]
            \draw[line width=1pt, fill=cyan, color=cyan] (0.75,4.25) .. controls (0.87,4.12)  and (1,3.9) .. (1,3.8)
            .. controls (1,3.75) and (0.8, 3.62) .. (0.8,3) -- (0.75, 1.8);
        \end{scope}
        
        \foreach \x in {.25,.75,...,4.75} {%
        \draw ($(\x,-3) + (0,-2pt)$) -- ($(\x,-3) + (0,2pt)$)
        node [above] {};
        }
        \draw(1.75,-3) node[above] {$x$};
        \draw(2.75,-3) node[above] {$y$};
        \draw(5,-3) node[right] {$\mu_G$};
        \draw(0,-5.5) node[left] {$\delta_z$};
    \end{scope}
    
\end{tikzpicture}
```

From this observation, we don't need to compute twice the distribution for $\mu = x$. If we compute all the $z_{t+1}$ distribution for each $\mu$ we expect, and compute $\mu$ vector as the climate variate, we can then *fill* the kernel.

# $\mu_G$ simulations

For this example we will use the species *Fagus abies* as an example.

```{r constant}
# Libraries
library(ggplot2)
library(dplyr)
library(matreex)

data("fit_Fagus_sylvatica")

# Load associated climate
data("climate_species")
climate <- subset(climate_species, sp == "Fagus_sylvatica", select = -sp)
```

## Get $\mu_G$ range

The first step consists of computing all $\mu_G$ values for a species in its climatic and basal area range. For this, we compute the $\mu_G$ range of the species and simulate a matrix for each values of $\mu_G$, with a small step. Keep in mind that the time of this computation rely mostly on the step and this step will influence the precision of the simulation later. The mu range step is done inside the `make_mu_gr` function call, but it's possible to check it before.

```{r range_mu}
mu_range <- getRangemu(
    climate = climate, fit = fit_Fagus_sylvatica, BA = seq(0, 200, by = 10), 
    mesh = seq(90, get_maxdbh(fit_Fagus_sylvatica) * 1.1, by = 2))
mu_range
```

*Note : Low values highly depend on maximum basal area, so, one can limit the $\mu_G$ computation time if he expect the basal area to stay low during simulation.*

Making the mu matrix takes more or less the same arguments as the function `make_IPM` in term of integration levels for Gauss-Legendre and mid-bin methods. The important argument that limit computation time is `stepMu` variable.

```{r make_mu}
mu_Fagus_sylvatica <- make_mu_gr(
    species = "Fagus_sylvatica", fit = fit_Fagus_sylvatica,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
    verbose = TRUE, stepMu = 0.1)
```

This object is a matrix with as many row as integrated cells for a given $\mu_G$, and a range of $\mu_G$ in column. During simulation, the loop will compute the $\mu_G$ needed for climate and basal area for a given time and extract from the matrix the required columns. 

**This step is 5 times longer than extracting the IPM matrix, this is the bottleneck during simulation.**

```{r get step_IPM}
step_climate <- subset(
    climate, 
    climate_label == "optimum", 
    select = -climate_label
)
step_climate <- drop(as.matrix(step_climate))
Fagus_IPM_BA20 <- get_step_IPM(x = mu_Fagus_sylvatica, BA = 20, 
                               climate = step_climate, sim_corr = "cut")
Fagus_IPM_BA20[12:17, 12:17]
dim(Fagus_IPM_BA20)
```

## Simulations

Once the $\mu_G$ matrix is created, creating species and simulating is really close to simulating with IPM. Because the $\mu_G$ matrices are not defined for a specific climate, we need to set one to simulate on it. For now we only set a single climate that will not change during simulation.

```{r simulation}
time <- 100

Fagus_sylvatica <- species(IPM = mu_Fagus_sylvatica, init_pop = def_initBA(20), 
                       harvest_fun = def_harv)

forest <- forest(species = list(mu_Fagus = Fagus_sylvatica))
set.seed(42)
memor_mu <- sim_deter_forest(forest, tlim = time, climate = step_climate,
                                    equil_dist = time, equil_time = time,
                                    verbose = TRUE, correction = "cut")

memor_mu %>%
    filter(var %in% c("BAsp", "H", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

We can compare this with a classical IPM simulation to see that it's not so different. Obviously this takes some times since integration takes longer. To limit this, we only integrate for basal area up to 40.

```{r compare_IPM}
ipm_Fagus <- make_IPM(
    "Fagus_sylvatica", step_climate, "opt_Fagus_clim", fit = fit_Fagus_sylvatica,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
    BA = 0:40, verbose = TRUE
)
Fagus_sylvatica_ipm <- species(IPM = ipm_Fagus, init_pop = def_initBA(20), 
                               harvest_fun = def_harv)
forest_ipm <- forest(species = list(ipm_Fagus = Fagus_sylvatica_ipm))
set.seed(42)
memor_ipm <- sim_deter_forest(forest_ipm, tlim = time, 
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut")
```

```{r compare_plot, fig.cap="comparison of simulations with integrated matrix and $\\mu_G$ approximation"}
e_memor <- dplyr::bind_rows(ipm = memor_ipm, mu = memor_mu, .id = "meth")
par(mfrow= c(1, 2))
e_memor %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = meth)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4, linetype = "dotted") + geom_point(size = .4)

e_memor %>%
    filter(var %in% c("BAsp"), ! equil) %>%
    group_by(time) %>% summarise(value = diff(value)) %>%
    ggplot(aes(x = time, y = value)) +
    ylab("BA difference") +
    geom_line(linewidth = .4, linetype = "dotted") + geom_point(size = .4)
```

## Climate evolution in time

Previous simulation specified a single climate that is used from start to finish, with only a variation on BA. The interest of using pre-computed values is too create new IPM matrix for a lot more climates. But as integrating an IPM takes around 2 or minutes, integrating a climate gradient would takes way too long, and only to use few basal area values when we integrate between 0 and 200.

Of course we now need to define a climate at each time step of the future simulation. For example, the simulation will start on optimum and migrate to hot edge in a hundred years, with a small random effect.

*It's possible to build this table by hand or provide a precise climate evolution table, the function shown is a quick example to add noise into a simulation*

```{r build_climate}
exp_clim <- clim_gradient(climate,  start_clim = "opti", end_clim = "hot",
                           times = time, noise = rnorm, sigma = c(100, 0.01), 
                          seed = 42)

par(mfrow= c(1, 2))
plot(1:time, exp_clim$sgdd, type = "l", ylab = "sgdd", xlab = "time")
plot(1:time, exp_clim$wai, type = "l", ylab = "wai", xlab = "time")
```

Once the climate is set, we can simulate a species. I recreate the species to start with a size distribution closer to the equilibrium reached in previous simulations.

```{r clim_sim}
distrib <- filter(memor_mu, equil, var == "n") %>% pull(value)
Fagus_sylvatica <- species(IPM = mu_Fagus_sylvatica, init_pop = def_init_k(distrib), 
                       harvest_fun = def_harv)
clim_forest <- forest(species = list(clim_Fagus = Fagus_sylvatica))
set.seed(42)
memor_clim <- sim_deter_forest(clim_forest, tlim = time, climate = exp_clim,
                                      equil_dist = time, equil_time = time,
                                      verbose = TRUE, correction = "cut")

memor_clim %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4) +
    NULL
```

## Testing effect of stepMu

The BA difference is already pretty low between IPM and $\mu_G$ simulations, but this section will illustrate that it's directly linked with `stepMu` value. Error is negligible for stepMu of 0.0001, but computing $\mu_G$ values takes around 2 minutes.


<details>
    <summary>Code to generate figure</summary>
```{r stepMu, eval = FALSE}
# Computation is long here
steps <- c(0.05, 0.01, 0.005, 0.001, 0.0005)
diff_list <- vector("list", length(steps))

for(s in seq_along(steps)){
    cat(sprintf(" %.4f", steps[s]))
    mu_Fagus <- make_mu_gr(
        species = "Fagus_sylvatica", fit = fit_Fagus_sylvatica,
        mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Fagus_sylvatica) * 1.1),
        stepMu = steps[s])
    
    Fagus <- species(IPM = mu_Fagus, init_pop = def_initBA(20), 
                     harvest_fun = def_harv)
    
    forest <- forest(species = list(Fagus = Fagus))
    set.seed(42)
    memor <- sim_deter_forest(forest, tlim = time, 
                                     climate = step_climate,
                                     equil_dist = time, equil_time = time,
                                     correction = "cut") 
    
    diff_list[[s]] <- dplyr::bind_rows(ipm = memor_ipm, mu = memor, 
                                       .id = "meth") %>%
        filter(var %in% c("BAsp"), ! equil) %>%
        group_by(time) %>% summarise(value = diff(value))
}

names(diff_list) <- c("a: 5e-2", "b: 1e-2", "c: 5e-3", "d: 1e-3", "e: 5e-4")
res <- do.call("bind_rows", c(diff_list, list(.id = "stepMu")))

p <- res %>%
    ggplot(aes(x = time, y = value, color = stepMu)) +
    ylab("BA difference") +
    geom_line(linewidth = .4)
# png("stepMu.png")
p
# dev.off()
summary <- res %>% group_by(stepMu) %>%
    summarise(min = min(value), max = max(value), d = max - min)
```
</details> 

```{r, echo = FALSE}
knitr::include_graphics(system.file(
        "extdata", "stepMu.png",
        package = "matreex",
        lib.loc = NULL,
        mustWork = FALSE
    ))
summary <- structure(list(
    stepMu = c("a: 5e-2", "b: 1e-2", "c: 5e-3", "d: 1e-3", "e: 5e-4"),
    min = c(-0.198040903223223, -0.0637381853746728, -0.0455626793410637, -0.0312318108955409, -0.0293529789309943), 
    max = c(0, 0, 0, 0, 0), 
    d = c(0.198040903223223, 0.0637381853746728, 0.0455626793410637, 0.0312318108955409, 0.0293529789309943)), 
    class = c("tbl_df", "tbl", "data.frame"), row.names = c(NA, -5L))
summary
```


## Multiple species simulations

We had a second species with $\mu_G$ values. This is just so we can have a distribution from species equilibrium at the climate given at very start of this document.

```{r Abies_alba}
data("fit_Abies_alba")
mu_Abies_alba <- make_mu_gr(
    species = "Abies_alba", fit = fit_Abies_alba,
    mesh = c(m = 700, L = 90, U = get_maxdbh(fit_Abies_alba) * 1.1),
    verbose = TRUE, stepMu = 0.001)

Abies_alba <- species(IPM = mu_Abies_alba, init_pop = def_initBA(20), 
                      harvest_fun = def_harv)

forest <- forest(species = list(mu_Abies = Abies_alba))
set.seed(42)
memor_abies <- sim_deter_forest(forest, tlim = time, climate = step_climate,
                                       equil_dist = time, equil_time = time,
                                       verbose = TRUE, correction = "cut")

distrib_abies <- filter(memor_abies, equil, var == "n") %>% pull(value) 
Abies_alba <- species(IPM = mu_Abies_alba, 
                      init_pop = def_init_k(distrib_abies), 
                      harvest_fun = def_harv)
```

We create a forest with 2 species integrated with $\mu_G$ values. From this, we just need to run the simulation. Please note that the climate can be modified along time as before because all species use $\mu_G$ values. Another thing to note is the simulation time that is correlated with the number of $\mu_G$ species used.

```{r}
nsp_forest <- forest(species = list(Fagus = Fagus_sylvatica, Abies = Abies_alba))
set.seed(42)
memor_nsp <- sim_deter_forest(nsp_forest, tlim = time, climate = exp_clim,
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut") 

memor_nsp %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```

If we want to simulate with species defined from $\mu_G$ and IPM, we can mix them but the climate used to integrate the IPM will be used. Providing a climate is thus useless in this case. 

```{r}
Fagus_sylvatica_ipm$init_pop <- def_init_k(distrib)
msp_forest <- forest(species = list(Fagus = Fagus_sylvatica_ipm, Abies = Abies_alba))
set.seed(42)
memor_msp <- sim_deter_forest(msp_forest, tlim = time, climate = exp_clim,
                                     equil_dist = time, equil_time = time,
                                     verbose = TRUE, correction = "cut")

memor_msp %>%
    filter(var %in% c("BAsp", "N"), ! equil, value != 0) %>%
    ggplot(aes(x = time, y = value, color = species)) +
    facet_wrap(~ var, scales = "free_y") +
    geom_line(linewidth = .4) + geom_point(size = .4)
```
